On Extending a Full-Sharing Multithreaded Tabling Design with Batched Scheduling

### 2015-05-26 paper.tex
paper -> revising 1

Tabling~\cite{Chen-96} is an implementation technique that overcomes
some limitations of traditional Prolog systems in dealing with
redundant sub-computations and recursion. Tabling consists of storing
intermediate answers for subgoals in a proper data structure, called
the \emph{table space}, so that they can be reused when a repeated
subgoal appears during the resolution process. Tabling has become a
popular and successful technique thanks to the ground-breaking work in
the XSB Prolog system and in particular in the SLG-WAM
engine~\cite{Sagonas-98}, the most successful engine of XSB. The
success of SLG-WAM led to several alternative implementations that
differ in the execution rule, in the data-structures used to implement
tabling, and in the changes to the underlying Prolog
engine. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

@@

With the FS design, all tables are shared. Thus, since several threads
can be inserting answers in the same table, when an answer already
exists, it is not possible to determine if the answer is new or
repeated for a certain thread without further support. However, for
local scheduling this is not a problem because, for repeated and new
answers, local scheduling always fails. The problem is with batched
scheduling that requires further support to work with the FS design
since, with batched scheduling, only repeated answers should fail. We
have thus to detect when a newly found answer is also a new answer for
the thread at hand in order to ensure that new answers for a given
subgoal call are correctly propagated one and only once to all
thread's repeated calls.

@@

We now present experimental results about the usage of PAC in the FS
design with batched scheduling. Since without PAC the FS design would
not be able to be used with batched scheduling, to put results in
perspective, we will be showing also the results for local scheduling
and for the NS design. The environment for our experiments was a
machine with 32-Core AMD Opteron (TM) Processor 6274 (2 sockets with
16 cores each) with 32G of main memory, running the Linux kernel is
the 3.16.7-200.fc20.x86\_64 with Yap Prolog 6.3. For the experiments,
we used the \emph{TabMalloc} memory allocator~\cite{Areias-12b} and
the set of benchmarks described in~\cite{Areias-12b}. These benchmarks
create \emph{worst case scenarios}, where we are able to show the
lowest bounds of performance that each design might achieve when
applied/used in other real world applications/programs.

@@

By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
thread, on the minimum overhead ratio, local scheduling is better on
the NS design, while batched scheduling is better in the FS
design. For the average and maximum overhead ratios, in the FS design,
the best strategy is always the local scheduling. Additionally, the
difference between local and batched scheduling in the FS is higher
than in the NS design, which indicates the overhead that the PAC
strategy brings to the FS design. Recall, whenever an answer is found
during the evaluation, PAC requires threads to traverse their private
consumer data structures to check if the answer was already
propagated. This is the reason for difference in the overheads that we
are observing in the experimental results. As we scale the number of
threads, for the NS design both scheduling strategies have similar
minimum, average and maximum overhead ratios. For the FS design, the
best minimum overhead ratio is always for the batched scheduling. For
the average and maximum overhead ratio, local scheduling is always
better than batched scheduling. However, for the average ratio, the
results between both strategies is quite close. For the maximum
overhead ratio, the difference between both strategies is
significantly higher than in the NS design. In conclusion,
experimental results show that on average, the PAC strategy does not
seem to have a big impact in the performance, however it still leaves
room for further improvements. The difference between both local and
batched scheduling is significantly higher in the FS design than in
the NS design.

>>>>

Tabling~\cite{Chen-96} is an implementation technique that overcomes
some limitations of traditional Prolog systems in dealing with
redundant sub-computations and recursion. Tabling consists in storing
intermediate answers for subgoals in a proper data structure, called
the \emph{table space}, so that they can be reused when a repeated
subgoal appears during the resolution process. Tabling has become a
popular and successful technique thanks to the ground-breaking work in
the XSB Prolog system and in particular in the SLG-WAM
engine~\cite{Sagonas-98}, the most successful engine of XSB. The
success of SLG-WAM led to several alternative implementations that
differ in the execution rule, in the data-structures used to implement
tabling, and in the changes to the underlying Prolog
engine. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

@@

With the FS design, all tables are shared. Thus, since several threads
can be inserting answers in the same table, when an answer already
exists, it is not possible to determine if the answer is new or
repeated for a certain thread without further support. However, for
local scheduling this is not a problem because, for repeated and new
answers, local scheduling always fails. The problem is with batched
scheduling that requires further support to work with the FS design
since, with batched scheduling, only repeated answers should fail. We
have thus to detect when a newly found answer is also a new answer for
the thread at hand in order to ensure that new answers for a given
subgoal call are correctly propagated once and only once to all
thread's repeated calls.

@@

We now present experimental results about the usage of PAC in the FS
design with batched scheduling. Since without PAC the FS design would
not be able to be used with batched scheduling, to put results in
perspective, we will be showing also the results for local scheduling
and for the NS design. The environment for our experiments was a
machine with 32-Core AMD Opteron (TM) Processor 6274 (2 sockets with
16 cores each) with 32G of main memory, running the Linux kernel
3.16.7-200.fc20.x86\_64 with Yap Prolog 6.3. For the experiments, we
used the \emph{TabMalloc} memory allocator~\cite{Areias-12b} and the
set of benchmarks described in~\cite{Areias-12b}. These benchmarks
create \emph{worst case scenarios}, where we are able to show the
lowest bounds of performance that each design might achieve when
applied/used in other real world applications/programs.

@@

By observing Table~\ref{tab_batched_overhead}, we can see that, for 1
thread, on the minimum overhead ratio, local scheduling is better on
the NS design, while batched scheduling is better in the FS
design. For the average and maximum overhead ratios, in the FS design,
the best strategy is always the local scheduling. Additionally, the
difference between local and batched scheduling in the FS is higher
than in the NS design, which indicates the overhead that the PAC
strategy brings to the FS design. Recall, whenever an answer is found
during the evaluation, PAC requires threads to traverse their private
consumer data structures to check if the answer was already
propagated. This is the reason for difference in the overheads that we
are observing in the experimental results. As we scale the number of
threads, for the NS design both scheduling strategies have similar
minimum, average and maximum overhead ratios. For the FS design, the
best minimum overhead ratio is always for the batched scheduling. For
the average and maximum overhead ratio, local scheduling is always
better than batched scheduling. However, for the average ratio, the
results between both strategies are quite close. For the maximum
overhead ratio, the difference between both strategies is
significantly higher than in the NS design. In conclusion,
experimental results show that on average, the PAC strategy does not
seem to have a big impact in the performance, however it still leaves
room for further improvements. The difference between both local and
batched scheduling is significantly higher in the FS design than in
the NS design.
